---
title: "ESM 211: adehabitatHR"
subtitle: "Modeling California gnatcatcher home ranges from 1990 and 2024"
author: "FIRST LAST"
date: February 13, 2025
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Overview

For this assignment, you will use the `adehabitatHR` package in R to model home ranges for California gnatcatcher (*Polioptila californica*, CAGN) - a bird that is native to California and Mexico. Species occurence data was downloaded from the Global Biodiversity Information Facility (GBIF) database from 1990 to present day. This data is saved in a CSV file called `p_californica`. Using a subset of this data, an example is provided for you that models the CAGN home range for 1990 to 1992 and calculates the home range size. After working through the 1990-1992 example, it will then be up to you to conduct the same analysis for a subset of the data for the year 2024 to compare how the species home range has changed over time.

There are seven main steps to follow:

1.  Load and filter data

2.  Create Spatial Points Data Frame

3.  Visualize Data

4.  Kernel Density Estimation

5.  Create Spatial Polygon Data Frame

6.  Better Visualization of Data

7.  Calculate Home Range Area

```{r}
# clear environment 
rm(list = ls())
```

## Load Libraries

```{r}
# Just in case you don't have these installed
# install.packages("tidyverse") 
# install.packages("here")  
# install.packages("janitor") 
# install.packages("adehabitatHR")  
# install.packages("leaflet") 
# install.packages("sf")  

#load libraries
library(tidyverse)    #for tidy data
library(here)         #for loading data
library(janitor)      #for cleaning data
library(adehabitatHR) #for modeling home range
library(leaflet)      #for mapping
library(sf)           #for mapping
```

## Example: 1990-1992 CAGN Home Range

### Load and Filter Data

The first thing we need to do is bring in the California gnatcatcher (CAGN) observation data and filter for observations from 1990 to 1992. As mentioned above, we've provided an example using this subset of the data below. You will then reproduce this workflow with the 2024 observations to determine changes in CAGN home range size.

```{r}
# Load gnatcatcher occurrence data downloaded from GBIF; note that "data" refers to an optional folder within your working directory and can be removed if appropriate
gnat <- read.csv(here("data", "p_californica.csv"))
```

```{r}
#create a subset of data for 1990 to 1992
gnat90_92 <- gnat |> 
  filter(year %in% 1990:1992)
```

### Create Spatial Points Data Frame

Once you've read in the data and filtered for 1990 to 1992 observations, we need to create a Spatial Points Data Frame (SPDF) from our latitude and longitude coordinates and specify the Coordinate Reference System (CRS) for mapping purposes. We do this with the `SpatialPointsDataFrame` function which requires several arguments, `coords` is the data frame containing the lat/long coordinates, `data` to define the data set, and `proj4string` to define the desired CRS projection. Be sure to check the CRS of your new object, `spdf90_92`.

```{r}
#Create a variable "prj" for the WGS 1984 CRS projection 
prj <- "+init=epsg:4326"

#Create a SpatialPointsDataFrame (SPDF) for the 1990-1992 gnatcatcher data
spdf90_92 <- SpatialPointsDataFrame(
             # Combine longitude and latitude into coordinate pairs
             coords = coordinates(cbind(gnat90_92$decimalLongitude,
                                        gnat90_92$decimalLatitude)),
             # Use the original dataset to include additional attributes
             data = gnat90_92, 
             # Assign the defined WGS 1984 projection
             proj4string = CRS(prj))
```

```{r}
#double check CRS
st_crs(spdf90_92)
```

### Visualize Data

Now we will plot a quick visualization of the SPDF from 1990-1992, using the function `plot`, to see what we are working with. It's not very exciting...yet.

```{r}
#make a quick plot
#pch and cex adjust the symbol and size
plot(spdf90_92, pch = 19, cex = .5)
```

### Kernel Density Estimation

Now, we will assess the species home range using Kernel Density estimation and visualize the output. The function `kernelUD` computes a Kernel Utilization Distribution (UD), which estimates the probability distribution of the species' locations based on observed data. The function requires two arguments: `xy` inputs the SPDF, and setting the `h` argument to "href" will automatically determine an optimal bandwidth for smoothing the density estimation using the reference bandwidth method. The function `image` allows a graphical display of the kernel density estimate.

```{r}
#Run the kernel density
kd90_92 <- kernelUD(xy = spdf90_92, #Input SPDF
                    h = "href")     #"href" stands for "reference bandwidth", which is a method for estimating the smoothing parameter, h, based on the data.


#Visualize the kernel density estimation as a raster-like image
image(kd90_92)
```

### Create Spatial Polygons from KDE

We now want to extract the gnatcatcher home range contours that we generated above. We will use the function `getverticeshr`, which calls for two arguments: `x` is the output from the kernel UD and `percent` indicates the percentage level for home-range estimation. We will extract home range polygons for 75% and 90% levels and confirm that the output is a Spatial Polygons Data Frame. We can create a quick plot to visualize the 75% and 90% gnatcatcher home ranges.

```{r}
# Extract 75% and 90% home range polygons
hr75_9092 <- getverticeshr(x = kd90_92, 
                           percent = 75)

hr90_9092 <- getverticeshr(x = kd90_92, 
                           percent = 90)

# Check the class of the extracted object (should be SpatialPolygonsDataFrame)
class(hr90_9092)
```

```{r}
#quick plot of 90% and 75% Kernel Density home ranges
plot(hr90_9092, col = "purple", main = "1990-1992 Kernel Density Home Ranges")
plot(hr75_9092, add = TRUE, border = "red")  # Add 75% range in red
```

### Better Visualization Methods

Now that we have our kernel density home ranges we can make some prettier visualizations using `ggplot` and `leaflet`. First, we will need to convert our spatial polygons into a simple feature (sf) format - the necessary format to use `ggplot` and `leaflet`. We will use the function `st_as_sf` to convert our spatial polygons into an sf object.

#### ggplot

```{r}
# Convert to sf format for ggplot and leaflet
hr75_9092_sf <- st_as_sf(hr75_9092)
hr90_9092_sf <- st_as_sf(hr90_9092)
```

```{r}
# Plot with ggplot
ggplot() +
  #Add the 90% Kernel density home range 
  geom_sf(data = hr90_9092_sf, 
          aes(fill = "90% Home Range"),  # Assign label for legend
          alpha = 0.4) +                 # Adjust transparency
  #Add the 75% Kernel density home range
  geom_sf(data = hr75_9092_sf, 
          aes(fill = "75% Home Range"),  
          alpha = 0.4) +
  #Add a legend 
  scale_fill_manual(name = "1990-1992 Home Range", 
                    values = c("90% Home Range" = "purple", 
                               "75% Home Range" = "red")) +
  theme_minimal()
```

#### leaflet

```{r}
# Leaflet map
leaflet() |>
  #Add a base map
  addTiles() |>
  #Add the 90% Kernel Density home range
  addPolygons(data = hr90_9092_sf, 
              color = "purple", 
              fillOpacity = 0.3) |>
  #Add the 75% Kernel Density home range
  addPolygons(data = hr75_9092_sf, 
              color = "red", 
              fillOpacity = 0.3) |> 
  #Add a legend
  addLegend(position = "bottomleft",
            colors = c("purple", "red"), 
            labels = c("90% kd", "75% kd"),
            title = "1990-1992 Home Range",
            opacity = 1)
```

### Calculate Home Range Area

Now we will calculate the area of the 75% and 90% home ranges using the `st_area` function. First we need to convert our simple features (sf) into a suitable coordinate reference system (CRS) that preserves distances and area. We will use the function `st_transform` to convert to the UTM Zone 11 CRS. Then we calculate the home range areas and convert to kilometer squared to compare the size differences.

```{r}
# Calculate the area of the home ranges

# Ensure sf is suitable CRS to preserve distances and area (e.g., UTM Zone 11)
hr90_9092_sf <- st_transform(hr90_9092_sf, crs = 32611)  
hr75_9092_sf <- st_transform(hr75_9092_sf, crs = 32611)  

# Calculate area in sq. kilometers by dividing by 1,000,000 b/c UTM is sq. meters
area_hr90_9092 <- as.numeric(st_area(hr90_9092_sf)) / 1000000
area_hr90_9092  # 90% Total home range area

area_hr75_9092 <- as.numeric(st_area(hr75_9092_sf)) / 1000000
area_hr75_9092  # 75% Total home range area
```

You should get 325331.9 km\^2 for the 90% total home range area and 162603 km\^2 for the 75% total home range area. We expect the 75% home range to be smaller than the 90% home range because there is a smaller proportion of occurrences being used to calculate the KDE.

## Portfolio Assignment: 2024 CAGN Home Range

Now that you have successfully modeled and visualized the California gnatcatcher home ranges from 1990-1992, we want to determine how the species range has changed over time by modeling the 2024 home range! You can follow the same workflow as above but for the 2024 subset of the data.

### Filter Data

```{r}
#create a subset of data for 2024
gnat2024 <- gnat |> 
  filter(year == 2024)
```

### Create Spatial Points Data Frame

```{r}
#Create a variable "prj" for the WGS 1984 CRS projection 
prj <- "+init=epsg:4326"

#Create a SpatialPointsDataFrame (SPDF) for the 2024 gnatcatcher data
spdf2024 <- SpatialPointsDataFrame(
             # Combine longitude and latitude into coordinate pairs
             coords = coordinates(cbind(gnat2024$decimalLongitude,
                                        gnat2024$decimalLatitude)),
             # Use the original dataset to include additional attributes
             data = gnat2024, 
             # Assign the defined WGS 1984 projection
             proj4string = CRS(prj)) #we created a variable "prj" for the WGS 1984 CRS projection above on line 78

#double check CRS
st_crs(spdf2024)
```

### Visualize Data

```{r}
#make a quick plot
plot(spdf2024, 
     pch = 19, # symbol type
     cex = .5) # symbol size
```

### Kernel Density Estimation

```{r}
# run the kernel density
kd2024 <- kernelUD(xy = spdf2024, # input SPDF
                   h = "href") # use the 'href' method to estimate smoothing

#Visualize the kernel density estimation as a raster-like image
image(kd2024)
```

### Create Spatial Polygons from KDE

```{r}
# Extract 75% and 90% home range polygons
hr75_2024 <- getverticeshr(x = kd2024, 
                           percent = 75) # 75% home range

hr90_2024 <- getverticeshr(x = kd2024, 
                           percent = 90) # 90% home range

# Check the class of the extracted object
# it should be a SpatialPolygonsDataFrame
class(hr90_2024) 
```

```{r}
#quick plot of 90% and 75% Kernel Density home ranges
plot(hr90_2024, col = "green", main = "2024 Kernel Density Home Ranges")
plot(hr75_2024, add = TRUE, border = "black")
```

### Better Visualization Methods

#### ggplot

```{r}
# Convert to sf format for ggplot and leaflet
hr75_2024_sf <- st_as_sf(hr75_2024)
hr90_2024_sf <- st_as_sf(hr90_2024)
```

```{r}
# plot with ggplot
ggplot() +
  # add the 90% Kernel density home range 
  geom_sf(data = hr90_2024_sf, 
          aes(fill = "90% Home Range"),  # assign label for legend
          alpha = 0.4) +                 # adjust transparency
  # add the 75% Kernel density home range
  geom_sf(data = hr75_2024_sf, 
          aes(fill = "75% Home Range"),  
          alpha = 0.4) +
  # add a legend 
  scale_fill_manual(name = "2024 Home Range", 
                    values = c("90% Home Range" = "purple", 
                               "75% Home Range" = "red")) +
  theme_minimal()
```

#### leaflet

```{r}
# leaflet map
leaflet() |>
  # add a base map
  addTiles() |>
  # add the 90% Kernel Density home range
  addPolygons(data = hr90_2024_sf, 
              color = "purple", 
              fillOpacity = 0.3) |>
  # add the 75% Kernel Density home range
  addPolygons(data = hr75_2024_sf, 
              color = "red", 
              fillOpacity = 0.3) |> 
  # add a legend
  addLegend(position = "bottomleft",
            colors = c("purple", "red"), 
            labels = c("90% kd", "75% kd"),
            title = "2024 Home Range",
            opacity = 1)
```

### Calculate HR Area

```{r}
# ensure sf is suitable CRS to preserve distances and area
hr90_2024_sf <- st_transform(hr90_2024_sf, crs = 32611)  
hr75_2024_sf <- st_transform(hr75_2024_sf, crs = 32611)  

# Calculate area in sq. kilometers (remember that UTM is sq. meters)
# divide by 1,000,000 to convert to km^2
area_hr90_2024 <- as.numeric(st_area(hr90_2024_sf)) / 1000000
area_hr90_2024  # 90% Total home range area

area_hr75_2024 <- as.numeric(st_area(hr75_2024_sf)) / 1000000
area_hr75_2024  # 75% Total home range area
```

## Compare Across Years

Now that we have the kernel density home ranges for 1990-1992 and for 2024 let's see how the species home range has changed. We can visualize the home ranges together in `leaflet` and compare the different home range sizes.

```{r}
# Reproject back to WGS84 to preserve mapping for Leaflet
hr90_9092_sf <- st_transform(hr90_9092_sf, crs = 4326)
hr90_2024_sf <- st_transform(hr90_2024_sf, crs = 4326)

# make sure this worked; confirm the CRS is WGS 1984
st_crs(hr90_9092_sf)
```

```{r}
# Now create the Leaflet map
leaflet() |>
  addTiles() |>
  addPolygons(data = hr90_9092_sf, # 1990-1992 90% home range for CAGN
              color = "purple", 
              fillOpacity = 0.3) |> 
  addPolygons(data = hr90_2024_sf, # 2024 90% home range for CAGN
              color = "green", 
              fillOpacity = 0.3) |>
  addLegend(position = "bottomleft",
            colors = c("purple", "green"), 
            labels = c("1990-1992", "2024"),
            title = "90% Home Range",
            opacity = 1)
```

```{r}
# compare 1990-1992 and 2024 home range sizes (hint: no new calculations are required)
# visually, we can just look at the leaflet plot

# calculate how much the home range has changed
# subtract the 2024 area from the 1990-1992 area
area_diff <- area_hr90_9092 - area_hr90_2024
```

Visually, we can see that the 2024 home range for the California gnatcatcher (CAGN) is smaller than the 1990-1992 home range. It's 90% home range doesn't extend as far into Baja California and Baja California Sur as it did in the early 1990s.

We also quantified the difference in home range area between 1990-1992 and 2024, finding that it decreased by approximately **181,591.28 km^2.**

## Questions

**1.  Why do you need to use a Spatial Points Data Frame for this analysis?**

We need a Spatial Points Data Frame (SPDF) to represent the geographical locations of the California gnatcatcher observations. As mentioned in the lecture, `adehabitatHR` function require spatial data as inputs, and a SPDF stores spatial coordinates and attributes of real-world locations. This allows us to perform spatial analyses in the `adehabitatHR` package, such as kernel density estimation and home range modeling like we did in this assignment, on the species' occurrence data.

**2.  What is the difference between a 75% and 90% kernel density estimation?**

The difference between a 75% and 90% kernel density estimation lies in the area they represent. The 75% kernel density estimation represents a smaller area with a higher density of observations, while the 90% kernel density estimation represents a larger area with a lower density of observations (wider distribution). In other words, the 75% kernel density estimation is more conservative and represents a tighter home range estimate, while the 90% kernel density estimation is more expansive and includes a broader range of potential habitat.

**3.  How has the California gnatcatcher home range size changed over time?**

The California gnatcatcher home range size has decreased over time. Comparing the 1990-1992 home range to the 2024 home range, we observed visually and quantitatively that the 2024 home range is smaller. The 90% home range area decreased by approximately **181,591.28 km^2** from 1990-1992 to 2024, indicating a reduction in the species' spatial distribution.




