---
title: "`popbio` portfolio assignment"
format: html
editor: visual
---

## Load packages

```{r}
library(tidyverse)
library(popbio)
```

## Load the prepared hudsonia data

```{r}
hudsonia <- hudsonia
```

## Determine the population growth rate for each year of available data (1985-1988) using the deterministic approach

```{r}
# extract the data for each year
A85 <- hudsonia[["A85"]]
A86 <- hudsonia[["A86"]]
A87 <- hudsonia[["A87"]]
A88 <- hudsonia[["A88"]]

# determine the population growth rate for each year
gr85 <- lambda(A85) # 0.9593
gr86 <- lambda(A86) # 1.0098
gr87 <- lambda(A87) # 0.8453
gr88 <- lambda(A88) # 1.0183
```

The population growth rates for the years 1985 to 1988 are as follows:

- **1985**: 0.9593
- **1986**: 1.0098
- **1987**: 0.8453
- **1988**: 1.0183

## Determine the stochastic growth rate

```{r}
# compile all projection matrices into one list
matrices <- list(A85, A86, A87, A88)
names(matrices) <- c("A85", "A86", "A87", "A88")

# find stochastic growth rate for 50,000 iterations
sgr_hudsonia <- stoch.growth.rate(matrices, verbose = FALSE)
sgr_hudsonia_exp <- exp(sgr_hudsonia$approx) # 0.9636
```

The stochastic growth rate for the years 1985 to 1988 is approximately **0.9636**.

## Question: Compare the growth rates calculated using the two methods. How are they different? Why?










